|

### 2023-09-28 src/manuscript.qmd
fix some citations

---
title: |
  Economic concerns appear to be weak predictors of White political identity
short-title: Economic concerns
published: "Manuscript draft: please do not share without author permission."
code-repo: "For replication, go to: <https:://github.com/DamonCharlesRoberts/White-identity-sources>."
author:
  - name: Damon C. Roberts
    email: damon.roberts-1@colorado.edu
    orcid: 0000-0002-4360-3675
    title: PhD Candidate
    affiliations:
      - id: CU
        name: University of Colorado Boulder
        department: Political Science
        address: 333 UCB
        city: Boulder
        region: CO 
        postal-code: 80309-0333
abstract: |
    Do economic or political threats explain reported White Political Identity? Overall, the social identity literature would suggest that White Political Identity would increase in response to economic threats. However, a number of those that study White Political Identity, specifically, argue that it results from concern about political influence. Considering what Whiteness means historically and contemporaneously, I argue that we should expect that political threats reflect stronger associations with White Political Identity. Using data from the 2012, 2016, and 2020 American National Election Study, I consider a single model using penalized regression containing proxies of economic and political threats. I find evidence suggesting that in the post-Trump era, White political identity is strongly associated with reported feelings of Whites’ loss of political influence as opposed to economic threats, as some suggest and may expect.
thanks: |
    Word count (including references): 3898. I would like to thank Jennifer Wolak for her feedback on the many drafts of this manuscript. I would also like to thank the Keller Center for the First Amendment for the research appointment that gave me the space to work on the early stages of this project. I would also like to thank Anand Sokhey and John Griffin for their advice and suggestions in the early drafts of this paper. I would also like to thank Tabitha Bonilla, Nicole Yadon, and participants and audience members of a Midwestern Political Science Association panel for which an earlier draft this idea was presented. I would also like to thank Tamar Malloy, Andrew Philips, Kathryn Schauer, Rachel O'Neal, Alexander Jensen, Komal Kaur, Andrew Engelhardt, and Madeline Mader for their encouragement and feedback at various stages of this project's development. All errors contained in this manuscript are ultimately mine.
additional-info: |
  I have no known conflict of interest to disclose.
keywords:
  - White Political Identity
  - economic threat
  - cultural threat
date: today
bibliography: "../assets/references.bib"
format:
  hikmah-pdf:
    cite-method: citeproc
    latex-output-dir: ../out/
    geometry:
      - top=1in
      - bottom=1in
      - left=1in
      - right=1in
      - heightrounded
    linestretch: 2
    fontsize: 12pt
    appendix-style: default
    include-in-header:
      text: |
        \usepackage{pdflscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
linestretch: 2
execute:
  echo: false
  warning: false
  message: false
params:
  replicate: false
---

@@

One perspective on what explains racial identity among Whites in the United States are that they are responding to threats to their economic status and power. Realistic Conflict Theory suggests that a racial identity and racial prejudice come as a response to a collective concern among members of a particular racial group about the loss of economic influence and the status a group holds relative to other racial groups [@blumer_race_1958;@bobo_perceptions_1996]. Some evidence suggests that Whites living in areas with shrinking job opportunities respond to this with an increased desire to protect the interests of their racial confederates [@baccini_gone_2021]. Others have experimental evidence supporting this perspective that White Political Identity is motivated by material concerns [@perez_politics_2022]. Further, @jardina_White_2019 who provides one of the most comprehensive treatments of conceptualizing, measuring, and examining the outcomes of White Political Identity includes indicators of economic status and power in the primary model.

Social identity theory, and its offshoots, would suggest an alternative perspective: that Whites expressing a racial identity are responding to political, cultural, and societal threats. While these threats are a bit more hard to define, there are a number of examples that demonstrate Whites are responsive to cultural or political changes that present as a threat. A common touchstone of Social Dominance Theory suggests that members of a racial group with the highest status are likely to find ways to maintain the racial hierarchy [@lange_social_2012]. From this perspective, many argue that Whites in the United States have and continue focus their efforts on maintaining influence in our institutions so that they may continue to wield political influence [@rile_hayward_how_2013;@lopez_White_2006]. For example, @mutz_status_2018 finds little support for the claim that the 2016 Presidential Election -- which support for Trump is associated with an expression of racial identity among Whites [@jardina_disavowing_2021] -- was a response to the economic effects of a globalizing economy and instead argues that it was a response to concerns about the cultural changes brought about by globalization and the racial-diversity it promotes. Others demonstrate that the behavioral manifestations of a White constituency in the United States was a reflection of the saliency of race in politics [@agadjanian_changing_2021].

@@

In this note, I am using a popular conceptualization of racial identity among Whites: that expression of it reflects homophily toward's one's own racial group [@ellemers_self_2002;@jardina_White_2019]. As identity conceptualized this way may predict negative intergroup attitudes, it reflects a desire to find ways to find similarities rather than differences between one's self and other in-group members [@ellemers_self_2002]. Identity is largely an elusive concept to operationalize [see @leach_group_2008]. Though there are implicit measures of White Political Identity [see @knowles_White_2005], the common and measure of White Political Identity is a response to the question, "How important is being White to your identity?" This measure captures a significant feature of an identity: the centrality or the importance of that identity to an individual [@jardina_White_2019]. Though a rather rough measure of something as complex as identity, it has been shown to produce similar results to more complex measures of White Political Identity and holds a significant amount of face validity [@jardina_White_2019].

@@

Here I conceptualize of political threat as perceptions that politics and political institutions no longer primarily serves White people. Existing work demonstrates that Whites who express a desire to maintain the racial status quo with Whites on top are more likely to express the belief that Whites are discriminated against at similar levels as Blacks [@marshburn_dominance_2022]. As Whiteness is often a central component of Whites' conceptualization of American society [see @green_migration_2013], changes to "traditional American values" may reflect concern about changes to the influences of White values. Finally, a common finding is that expression of White Political Identity is heavily wrapped up in the perception that racial stratifications are by meritocratic means [@lowery_paying_2012]. As members of a dominant social group, when given information that either threatens one's belief that their merits are earned or that their group operates as a way to maintain inequality, we expect lower levels of expression in identification with that group. The classic items of racial resentment construct capture these ideas in that they revolve around the idea that Blacks are not disadvantaged as the result of systematic oppression by Whites. Though many have often used racial resentment as a measure of prejudice which would be a possible outcome of White Political Identity, some make the compelling argument that it is distinct as a concept and operationalization [@davis_racial_2022].

@@

Racial groups in the U.S. contain a number of these features. As a result, a burgeoning literature in political science that has made this connection takes the racial identity of Whites more seriously as one that informs their political attitudes [see @jardina_White_2019]. Race in America is subjectively defined [see @rile-hayward_2013], and these definitions are deeply embedded and reinforced by our institutions [@lopez_2006]. Whiteness and one's willingness to identify with that group is important [@wong_cho_2005] as it reflects a desire for an association with the history and affordances of being White [@mcdermott_samson_2005]. 
->
Recent work makes the case that we should seriously consider Whites' racial identity as an identity [@jardina_White_2019]. As an identity, much of the literature view these attitudes resulting from Whites' racial identity as stemming from economic threat. Some studying White political identity find economic predictors as substantively and statistically significant [see @perez_et-al_2022_pp;@lindsay_relationship_2023 on their racial attitudes writ large]. Observational evidence supports this claim as well. For example, Trump's messaging about job market competition with immigrants appeared to be successful at attracting White voters [@bunyasi_2019]. After all, the identity literature, more broadly, often focuses on economic threats. Evolutionary psychologists often take the position that identity and the resulting attitudes and behaviors directed toward out-groups are the result of competition over material resources [see @huddy_2001]. Many political scientists use this as an explanation for negative attitudes directed toward immigrants among those who feel insecure about their job prospects [@pardos-prado_xena_2019]. Others argue that deindustrialization and job loss activate Whites' concerns about the racial hierarchy and predict their support for Republican challengers that express support for keeping the racial status quo [@baccini_weymouth_2021_apsr].

As much of the literature on Whites' racial attitudes rely heavily on social identity theory as a paradigm, a significant portion of it presents the source of these group-based views as responses to economic threat. However, some disagree with this position [see @jardina_White_2019] and view these attitudes as the result of changes to the racial hierarchy. These other perspectives provide explanations for Whites' attitudes and conceptualize the definition of Whiteness and White Political Identity as a desire to retain political power.

@@

While a critical component of identity is a feeling of your fate as linked with others in your group [@leach_et-al_2008], this linked fate is the connection between attitudes and material forms of threat but that identity, in and of itself, is better conceptualized as a result of cultural or ideational threat [@donnelly_2020]. Aligning with the contextual account of racial identity for Whites, many conceptualize White Political Identity as an expressed desire for a recognition of privilege and status by institutions [@lopez_2006] and a connection with the history that connects Whites to their status [@mcdermott_samson_2005]. The prominent account of White political identity and its implications in contemporary American politics argues for a strong connection between White political identity and a desire to protect this racial hierarchy and status. This conceptualization reflects the tendency to report in-group-oriented, not necessarily out-group-oriented, attitudes [@jardina_White_2019].

@@

To account for potential problems stemming from missing data, I impute my data by using multiple imputation with chained equations (MICE) where the particular model I use is a random forest model to allow for more flexibility and to reduce model dependence [@buuren_mice_2011; @marbach_choosing_2022].^[I impute 10 datasets and simultaneously fit my Bayesian LASSO on each dataset and pool the results. Uncertainty reflected in my reported credible intervals not only reflects the uncertainty from one model, but my reported uncertainty also reflects the uncertainty generated from the imputation procedure. One drawback with this particular approach is that my normalized split-$\hat{R}$ will likely be greater than the widely recognized 1.01 cutoff that indicates model convergence due to the pooling of my multiply imputed datasets.] I report the results of the model fitted using listwise deletion in the Supplementary Information.

>>>>

---
title: |
  Economic concerns appear to be weak predictors of White political identity
short-title: Economic concerns
published: "Manuscript draft: please do not share without author permission."
code-repo: "For replication, go to: <https:://github.com/DamonCharlesRoberts/White-identity-sources>."
author:
  - name: Damon C. Roberts
    email: damon.roberts-1@colorado.edu
    orcid: 0000-0002-4360-3675
    title: PhD Candidate
    affiliations:
      - id: CU
        name: University of Colorado Boulder
        department: Political Science
        address: 333 UCB
        city: Boulder
        region: CO 
        postal-code: 80309-0333
abstract: |
    Do economic or political threats explain reported White Political Identity? Overall, the social identity literature would suggest that White Political Identity would increase in response to economic threats. However, a number of those that study White Political Identity, specifically, argue that it results from concern about political influence. Considering what Whiteness means historically and contemporaneously, I argue that we should expect that political threats reflect stronger associations with White Political Identity. Using data from the 2012, 2016, and 2020 American National Election Study, I consider a single model using penalized regression containing proxies of economic and political threats. I find evidence suggesting that in the post-Trump era, White political identity is strongly associated with reported feelings of Whites’ loss of political influence as opposed to economic threats, as some suggest and may expect.
thanks: |
    Word count (including references): 3233 (4049). I would like to thank Jennifer Wolak for her feedback on the many drafts of this manuscript. I would also like to thank the Keller Center for the First Amendment for the research appointment that gave me the space to work on the early stages of this project. I would also like to thank Anand Sokhey and John Griffin for their advice and suggestions in the early drafts of this paper. I would also like to thank Tabitha Bonilla, Nicole Yadon, and participants and audience members of a Midwestern Political Science Association panel for which an earlier draft this idea was presented. I would also like to thank Tamar Malloy, Andrew Philips, Kathryn Schauer, Rachel O'Neal, Alexander Jensen, Komal Kaur, Andrew Engelhardt, and Madeline Mader for their encouragement and feedback at various stages of this project's development. All errors contained in this manuscript are ultimately mine.
additional-info: |
  I have no known conflict of interest to disclose.
keywords:
  - White Political Identity
  - economic threat
  - cultural threat
date: today
bibliography: "../assets/references.bib"
format:
  hikmah-pdf:
    cite-method: citeproc
    latex-output-dir: ../out/
    geometry:
      - top=1in
      - bottom=1in
      - left=1in
      - right=1in
      - heightrounded
    linestretch: 2
    fontsize: 12pt
    appendix-style: default
    include-in-header:
      text: |
        \usepackage{pdflscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
    #citeproc: false
    #filters: 
    #  - "_extensions/andrewheiss/wordcount/citeproc.lua"
    #  - "_extensions/andrewheiss/wordcount/wordcount.lua"
linestretch: 2
execute:
  echo: false
  warning: false
  message: false
params:
  replicate: false
---

@@

One perspective on what explains racial identity among Whites in the United States are that they are responding to threats to their economic status and power. Realistic Conflict Theory suggests that a racial identity and racial prejudice come as a response to a collective concern among members of a particular racial group about the loss of economic influence and the status a group holds relative to other racial groups [@blumer_race_1958;@bobo_perceptions_1996]. Some evidence suggests that Whites living in areas with shrinking job opportunities respond to this with an increased desire to protect the interests of their racial confederates [@baccini_gone_2021]. Others have experimental evidence supporting this perspective that White Political Identity is motivated by material concerns [@perez_politics_2022]. Further, @jardina_white_2019 who provides one of the most comprehensive treatments of conceptualizing, measuring, and examining the outcomes of White Political Identity includes indicators of economic status and power in the primary model.

Social identity theory, and its offshoots, would suggest an alternative perspective: that Whites expressing a racial identity are responding to political, cultural, and societal threats. While these threats are a bit more hard to define, there are a number of examples that demonstrate Whites are responsive to cultural or political changes that present as a threat. A common touchstone of Social Dominance Theory suggests that members of a racial group with the highest status are likely to find ways to maintain the racial hierarchy [@lange_social_2012]. From this perspective, many argue that Whites in the United States have and continue focus their efforts on maintaining influence in our institutions so that they may continue to wield political influence [@rile_hayward_how_2013;@lopez_white_2006]. For example, @mutz_status_2018 finds little support for the claim that the 2016 Presidential Election -- which support for Trump is associated with an expression of racial identity among Whites [@jardina_disavowing_2021] -- was a response to the economic effects of a globalizing economy and instead argues that it was a response to concerns about the cultural changes brought about by globalization and the racial-diversity it promotes. Others demonstrate that the behavioral manifestations of a White constituency in the United States was a reflection of the saliency of race in politics [@agadjanian_changing_2021]. Additional evidence suggests that rural Americans -- who acutely feel economically left-behind by their governments [@cramer_politics_2016] -- express more racial animus as opposed to White Political Identity as a story of economic threat may suggest [@dawkins_place_2023]. Rather non-urban Whites who express a White Political Identity are more likely to express an empathy for the rural White American who has been left-behind [@dawkins_place_2023].

@@

In this note, I am using a popular conceptualization of racial identity among Whites: that expression of it reflects homophily toward's one's own racial group [@ellemers_self_2002;@jardina_white_2019]. As identity conceptualized this way may predict negative intergroup attitudes, it reflects a desire to find ways to find similarities rather than differences between one's self and other in-group members [@ellemers_self_2002]. Identity is largely an elusive concept to operationalize [see @leach_group_2008]. Though there are implicit measures of White Political Identity [see @knowles_white_2005], the common and measure of White Political Identity is a response to the question, "How important is being White to your identity?" This measure captures a significant feature of an identity: the centrality or the importance of that identity to an individual [@jardina_white_2019]. Though a rather rough measure of something as complex as identity, it has been shown to produce similar results to more complex measures of White Political Identity and holds a significant amount of face validity [@jardina_white_2019].

@@

Here I conceptualize of political threat as perceptions that politics and political institutions no longer primarily serves White people. Existing work demonstrates that Whites who express a desire to maintain the racial status quo with Whites on top are more likely to express the belief that Whites are discriminated against at similar levels as Blacks [@marshburn_dominance_2022]. As Whiteness is often a central component of Whites' conceptualization of American society [see @green_migration_2013], changes to "traditional American values" may reflect concern about changes to the influences of White values. Finally, a common finding is that expression of White Political Identity is heavily wrapped up in the perception that racial stratifications are by meritocratic means [@lowery_paying_2012]. As members of a dominant social group, when given information that either threatens one's belief that their merits are earned or that their group operates as a way to maintain inequality, we expect lower levels of expression in identification with that group. The classic items of racial resentment construct capture these ideas in that they revolve around the idea that Blacks are not disadvantaged as the result of systematic oppression by Whites. Though many have often used racial resentment as a measure of prejudice which would be a possible outcome of White Political Identity, some make the compelling argument that it is distinct as a concept and operationalization [@darren_w_davis_racial_2022].

@@

Racial groups in the U.S. contain a number of these features. As a result, a burgeoning literature in political science that has made this connection takes the racial identity of Whites more seriously as one that informs their political attitudes [see @jardina_white_2019]. Race in America is subjectively defined [see @rile-hayward_2013], and these definitions are deeply embedded and reinforced by our institutions [@lopez_2006]. Whiteness and one's willingness to identify with that group is important [@wong_cho_2005] as it reflects a desire for an association with the history and affordances of being White [@mcdermott_samson_2005]. 
->
Recent work makes the case that we should seriously consider Whites' racial identity as an identity [@jardina_white_2019]. As an identity, much of the literature view these attitudes resulting from Whites' racial identity as stemming from economic threat. Some studying White political identity find economic predictors as substantively and statistically significant [see @perez_et-al_2022_pp;@lindsay_relationship_2023 on their racial attitudes writ large]. Observational evidence supports this claim as well. For example, Trump's messaging about job market competition with immigrants appeared to be successful at attracting White voters [@bunyasi_2019]. After all, the identity literature, more broadly, often focuses on economic threats. Evolutionary psychologists often take the position that identity and the resulting attitudes and behaviors directed toward out-groups are the result of competition over material resources [see @huddy_2001]. Many political scientists use this as an explanation for negative attitudes directed toward immigrants among those who feel insecure about their job prospects [@pardos-prado_xena_2019]. Others argue that deindustrialization and job loss activate Whites' concerns about the racial hierarchy and predict their support for Republican challengers that express support for keeping the racial status quo [@baccini_weymouth_2021_apsr].

As much of the literature on Whites' racial attitudes rely heavily on social identity theory as a paradigm, a significant portion of it presents the source of these group-based views as responses to economic threat. However, some disagree with this position [see @jardina_white_2019] and view these attitudes as the result of changes to the racial hierarchy. These other perspectives provide explanations for Whites' attitudes and conceptualize the definition of Whiteness and White Political Identity as a desire to retain political power.

@@

While a critical component of identity is a feeling of your fate as linked with others in your group [@leach_et-al_2008], this linked fate is the connection between attitudes and material forms of threat but that identity, in and of itself, is better conceptualized as a result of cultural or ideational threat [@donnelly_2020]. Aligning with the contextual account of racial identity for Whites, many conceptualize White Political Identity as an expressed desire for a recognition of privilege and status by institutions [@lopez_2006] and a connection with the history that connects Whites to their status [@mcdermott_samson_2005]. The prominent account of White political identity and its implications in contemporary American politics argues for a strong connection between White political identity and a desire to protect this racial hierarchy and status. This conceptualization reflects the tendency to report in-group-oriented, not necessarily out-group-oriented, attitudes [@jardina_white_2019].

@@

To account for potential problems stemming from missing data, I impute my data by using multiple imputation with chained equations (MICE) where the particular model I use is a random forest model to allow for more flexibility and to reduce model dependence [@buuren_mice_2011; @moritz_marbach_choosing_2022].^[I impute 10 datasets and simultaneously fit my Bayesian LASSO on each dataset and pool the results. Uncertainty reflected in my reported credible intervals not only reflects the uncertainty from one model, but my reported uncertainty also reflects the uncertainty generated from the imputation procedure. One drawback with this particular approach is that my normalized split-$\hat{R}$ will likely be greater than the widely recognized 1.01 cutoff that indicates model convergence due to the pooling of my multiply imputed datasets.] I report the results of the model fitted using listwise deletion in the Supplementary Information.
